Network Data Model and BerlinMOD Benchmark

### 2010-04-22 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
Updated

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{2.9.StableNewFlob}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past several data models for the representation of histories of  spatio-temporal data
objects have been developed. Among others we can categorize these data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. We selected two representants,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compared their capabilities with the \bmodb{}. In this paper we present our
translation from the \bmodb{} into the network constrained data model and show
that the network constrained data model outperforms the data model of free movement
in the two dimensional space by orders of magnitudes.

%The very good results of the network data model in this comparsion encourages us
%to propose a extension of the \bmodb{} with a additional query set BerlinMOD/Net.
%This new query set covers the specialised challenges for network constrained
%data models and enables us to use the \bmodb{} to compare the capabilities of
%different network data models with respect to this specialised network challenges.
\end{abstract}
\section{Introduction}
In the past several data models for the representation of spatial and
spatio-temporal data objects have been developed. Among others we can categorize
them into data models for objects moving freely in two dimensional space and
data models for network constrained moving objects. For both categories
several different data models have been presented like
\cite{335426,chenzaniolosqlst} for spatio-temporal data objects moving freely in
two dimensional space and \cite{1146465,956692,VazWolfNetMod} for
spatio-temporal data objects that are constrained by given networks,
to name just a few. Objects which are restricted to use existing networks, like
cars to road networks, can be represented as moving point object in both data
models, while objects, which are not restricted by a given network, like people,
can be represented as moving point object only in the data models of free movement
in space. So, why do we spend time in developing network data models, if the network
constrained moving objects can be represented elsewhere in the data model of free
movement in the space?

We decided to compare the capablities of the both different data models. For our
experiments we choosed one example data model as representative for
each of the both categories. Spatio-temporal objects moving freely in the two
dimensional space are represented by the data model presented in \cite{335426}.
And network constrained data models are represented by the network data model
presented in \cite{1146465}. We choosed this two representatives, because both
data models are available in the \secondo{} DBMS and use the same temporal
representation. So we can exclude that the use of different DBMS or temporal
representations bias the results of our comparision.

As we use \secondo{} it seems to be a logical next step to use the \bmodb{}
\cite{BerlinMODVLDB} to compare the capabilities of the
both data models. The \bmodb{} is best to our knowledge the first benchmark for
complete spatio-temporal database systems and it is also available in the \secondo{}
DBMS. Furthermore the \bmodb{} data model is the data model of freely movement
in two dimensional space that we use for our comparison. So we only have to
transform the \bmodb{} spatial and spatio-temporal data types once into our
network data model representation. This avoids sources of errors in transformation
and makes it more easy to control the query results for correctness. Furthermore
the data generated by the \bmodb{} data generator is restricted to the streets of the
German capital Berlin, such that it can be translated into a network data model
environment.

In this paper we describe the transformation of the \bmodb{} benchmark data and
queries into network data model representation. And show that the network data
model outperforms the data model of free movement in space significantly in our
experiments. The network data model uses less than 60\% of the storage space of
the data model of free movement in space, mostly caused by the fact that the number
of units for the moving point objects in the network data model is less than 50\% of
the number of units in the data model of free movement in space. The total run times
for the \bmodb{} of the network data model is less than 50\% of the total run time
of the data model of free movement in space. We think that this results show that
it is useful to develope specialised data models for network constrained moving objects
to save storage space and reduce query run times.

@@

The rest of the paper is organised as follows: In section \ref{sec:relWork} we give a
short reminder of the underlying \secondo{} DBMS (\ref{sec:secondo}), the \bmodb{}
(\ref{sec:bmodb}) and the two data models (\ref{sec:bmodbdatamod}
and \ref{sec:netdatamod}) we choosed as representatives for our comparision.
The translation of the \bmodb{} data and query set in the network data model
representation is described in section (\ref{sec:Translation}). The resulting
experimental benchmark setup is described in section \ref{sec:scenario} followed
by the results of our experiments in section (\ref{sec:results}).
%At least we present our proposed extension of the \bmodb{} for network data models in
%section \ref{sec:newqueries}.
We conclude our work in section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
In the past other data models for free movement in the space
\cite{335426,chenzaniolosqlst} and for network constrained movement have been
presented \cite{1146465,956692,VazWolfNetMod}. As well there are some more
benchmarks \cite{COSTBenchmark, QueriesTheodoridis} and database systems for
spatial and spatio-temporal datatypes \cite{HERMES,1054151}. But best to our
knowledge they don't actually provide a combination of different implemented
and supported data models together with a existing benchmark feasible for moving
objects in space and network constrained moving objects like the actual
\secondo{} DBMS. So it is self-evident for us to use the \secondo{} DBMS in
combination with the provided data models and the \bmodb{} to compare the
capabilities of the network constrained data model and the data model of free
movement in spache provided with the \secondo{} DBMS.

In the next subsections we give short reminders of the \secondo{} database system
(\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}), and the both data models
(\ref{sec:bmodbdatamod}, \ref{sec:netdatamod}) we used in our experiments.
\subsection{Secondo}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{686903,1054151} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce  new data types and operations on
this data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the user may publish his extensions
as \secondo{} plugin such that other users can use this plugin to extend their own
\secondo{} system to use the functionalities or repeat the published experiments.
\secondo{} is free available in the web \cite{secondoweb} and comes with a number
of already implemented spatial and spatio-temporal data types
and operations including the spatio-temporal data model of free movement in the
two dimensional space \ref{sec:bmodbdatamod} and the network data model
\ref{sec:netdatamod} used in our work. Furthermore the \bmodb{} described in
\ref{sec:bmodb} has been developed in the \secondo{} DBMS. For our experiments
we used the \secondo{} version \secver{}.
\subsection{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was presented in \cite{BerlinMODVLDB} \nocite{BerlinMOD} and the
provided scripts for the data generator are implemented as \secondo{} DBMS operations.
The \bmodb{} is available in the web \cite{berlinmodweb} and provides a well defined
data-set and queries for the experimental evaluation of the capabilities of
spatial and spatio-temporal database systems dealing with histories of moving
objects. The \bmodb{} emphasises the development of complete systems
and simplifies experimental repeatability pointing out the weakness and the potencies
of the benchmarked systems.

The data-sets of the \bmodb{} are created using the street map of the German
captial Berlin \cite{bbike} and statistical data about the regions of Berlin
\cite{bevberlin,berlinstadtatlas} as input relations.
The created moving objects represent cars driving in the streets of Berlin,
simulating the behaviour of people living and working in Berlin.
Every moving object has a home node and a work node. Every weekday each car will
do a trip from the home node to work the work node in the morning and vice versa
in the late afternoon. Besides this randomly choosen cars will make additional
trips in the evening and up to six times at the weekend to randomly choosen
targets in Berlin. The \bmodb{} uses the data model of free movement in two
dimensional space described in section \ref{sec:bmodbdatamod}. Because the \bmodb{}
generates all data sets restricted to the street map of Berlin the \bmodb{} can
also be used for network constrained data models, if the spatial and spatio-temporal
data types are translated into a corresponding network data model, like we did
for our experiments.

The number of observed cars and the duration of the observation period can be
influenced by the user setting the $SCALEFACTOR$ to different values in the data
generation script of the \bmodb{}. For example at $SCALEFACTOR$ 1.0 the data generator
creates 2000 moving point objects observed for 28 days. Each of them sending a
GPS-signal every 2 seconds. This simulated signals are simplified such that time
intervals when a car doesn't move or moves in the same direction at the same
speed are merged into one moving unit. This means that, if a car stays in front
of the work node for 8 hours there will be only one entry in the cars
history of movement with a time interval of 8 hours instead of 14.400 entries
one for each GPS interval.

@@

In the TBA we have two relations $dataMcar$ and $dataMtrip$. $dataMcar$ contains
the static data for each object like $licence$, $type$, and $model$ together with
an object identifier $moid$. $dataMtrip$ contains for each $moid$ several tuples
each of them containing all units of a single trip of the moving object or a
single unit for a longer stop. For example each time the car drives from home node
to work node is a single trip and each time the car parks in front of the office
is also a single trip.

Besides the moving objects the \bmodb{} provides several data sets each of them
containing 100 pseudo randomly generated data objects which are used in the
benchmark queries. Table \ref{tab:queryobjects} gives an overview of this query
objects.

\begin{table}[H]
  \begin{tabularx}{|l|X|}
    \hline
    \textbf{Name of Data Set}&\textbf{Tuple Content}\\
    \hline
    $QueryPoints$&Object identifier and \dt{point} values.\\
    \hline
    $QueryRegions$&Object identifier and \dt{region} value.\\
    \hline
    $QueryInstants$&Object identifier and time stamps.\\
    \hline
    $QueryPeriods$&Object identifier and space of time.\\
    \hline
    $QueryLicences$& Object identifier and a \dt{string} representing a licence value.\\
    \hline
  \end{tabularx}
  \caption{Query Object Relations of \bmodb{}}
  \label{tab:queryobjects}
\end{table}

@@

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item object identity (known / unknown),
  \item dimension (standard / spatial / temporal / spatio-temporal),
  \item query interval (point / range / unbounded),
  \item condition type (single object / object relations),
  \item aggregation (with or without aggregation).
\end{itemize}
We will present the 17 queries in more detail in section \ref{sec:queries}
together with our network data model translation of this queries.
\subsection{Data Model of Free Movement in 2D space}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the same data model of freely moving in
two dimensional space presented in \cite{594784,335426,352963}. Spatial
positions are assumed to be located into a two dimensional space. A single spatial
position is represented by the data type \dt{point}. A \dt{point} consists of a pair of
\dt{real} values interpreted as x,y-coordinates in the assumed two dimensional plane.

Streets are represented by \dt{line} values. A \dt{line} value consists of a set
of $half segments$ representing the geometry of the line in the two dimensional space.
Each $half segment$ consists of two \dt{point} values and a boolean value telling
if the left or the right point is the dominating point of the $half segment$.

@@

A car may be represented by a \dt{mpoint}. A \dt{mpoint} is a \dt{point} object changing
its value within time. Therefore a \dt{mpoint} consist of a set of units called
\dt{unit}(\dt{point}) (short form \dt{upoint}). Each \dt{upoint} consists of a time
interval and two \dt{point} values. The first \dt{point} value represents the
position of the \dt{mpoint} at the start of the time interval and the second
\dt{point} value represents the position of the \dt{mpoint} at the end of the
time interval. It is assumed that the point moves on the straight line between
this two points with constant speed, which is defined by the ratio from the
distance of the two points and the length of the time interval of the unit.
All units of a \dt{mpoint} must have disjoint time intervals, because a car
cannot be at two different positions at the same time.
The units are sorted by ascending time intervals.

@@

Some other data types of \secondo{} which are used in the \bmodb{} are shown in
table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{\textwidth}{|l|X|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type.\\
\hline
\dt{int} & Usual integer number.\\
\hline
\dt{real} & Usual real number.\\
\hline
\dt{instant} & A point in time.\\
\hline
\dt{periods} & A set of disjoint and not connected spaces of time.\\
\hline
\dt{mbool} & A time dependent boolean value, which will be constant \true{} or \false{}
within each \dt{ubool} \\
\hline
\dt{mreal} & Time dependent real number. Each unit will be defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\subsection{Network Data Model}
\label{sec:netdatamod}
The central idea of the network data model presented in \cite{1146465} is that
every movement is constrained by a given network and every position can be described
relative to this network. This make the data type \dt{network} to the central
data type in the network data model. All other data types of the network data model
are related to a given \dt{network} object by the unique network identifier that
is part of each \dt{network} object.

The \dt{network} object contains all spatial informations of the represented network
in three main relations, two sub relations and some indexes to support faster
access to the relations.

The relation $routes$ contains the attributes of the
streets like $id$, $route curve$, $route length$, and two boolean flags. The first
flag indicates if the route starts at the lexicographically smaller end point or not.
The second flag indicates if the lanes of the street are separated like on German
Highways or not.

The $junctions$ relation contains all attributes of the street
crossings like the two route identifiers of the first and second street meeting at the
crossing, the distance of the crossing from the start of the first respectively
second street, tuple identifiers of the both streets in the routes relation,
tuple identifiers of the sections connected by this junction in the sections relation,
and a connectivity code telling us which lanes of the two streets are connected
by the crossing.

The $sections$ relation of the \dt{network} object contains the
attributes of the street parts between two crossings or a crossing and
the end of the street. This are the route identifier of the street the section
belongs to, the tuple identifier of this street in the routes relation, the start
position and end position of the section on the street, the section curve, and
again two boolean flags with the same meaning as in the routes relation.

@@

The data type \dt{gpoint} represents single positions in a given network. Besides the
network identifier a \dt{gpoint} consists of a route identifier, a distance from
the start of the route to the position of the \dt{gpoint} and a $side$ value
({$up$, $down$, $none$) telling us if the position is reachable from the $up$
or the $down$ side of the route in case of seperated lanes. For simple
streets or positions which are reachable from both sides of the route the side
value is always $none$.

Parts of the network, regardless if they represent paths or regions, are given as
\dt{gline} values. Besides the network identifier a \dt{gline} consists of a set
of $route interval$s, and two boolean flags. The boolean flags tell us if the
\dt{gline} is defined and if the set of $route interval$s is sorted.

Each $route interval$ consists of a route identifier identifying the route
the route interval belongs to, and the start and the end position from the
route interval on this route
\footnote{In the original paper the $route interval$ includes a $side$ value
analogous to the \dt{gpoint}. But this parameter is not part of the implementation yet.}.
We call a \label{sec:sortedgline} set of $route interval$s sorted if the following
conditions are fullfilled:
\begin{itemize}
	\item all $route interval$s are disjoint
	\item the $route interval$s are stored in ascending order of their route identifiers
	\item if two disjoint $route interval$s have the same route identifier the
$route interval$ with the smaller start position is stored first
	\item for all $route interval$s the $startPosition \le endPosition$
\end{itemize}
Many algorithms take profit from sorted \dt{gline} values. For example: If $n$
is the number of the route intervals in the \dt{gline} the
decision, if a \dt{gpoint} is inside a \dt{gline} needs O($n$) time for unsorted
and O($\log n$) time for sorted \dt{gline} values,

Unfortunately not all \dt{gline} values can be stored sorted. If a \dt{gline}
value represents a path between two \dt{gpoint} in the network, we need the
route intervals exactly in the sequence they are used in the path. This will
nearly never be a sorted set like defined before. We solved this dilemma by
introducing the boolean sorted flag. Every algorithm which can take profit from
a sorted \dt{gline} checks this flag and uses the corresponding code. We store \dt{gline}
values sorted whenever this is possible to support faster query execution.

@@

In deviation from the original network data model we extended the implementation
of the \dt{mgpoint} with four additional attributes to support faster query execution:
\begin{enumerate}
	\item The total driven distance
	\item A sorted set of $route interval$s representing the positions ever
traversed by the \dt{mgpoint}
	\item A boolean defined flag for the set of route intervals
	\item A spatio-temporal minimum bounding box
\end{enumerate}
The sorted set of $route interval$s was introduced, because it makes it much
faster to decide if a \dt{mgpoint} ever passed a given network position or not.
Instead of a linear check of all $m$ \dt{ugpoint}s of a \dt{mgpoint} we can
perform a binary scan on the much lower number $r$ of the passed $route interval$s.
This reduces the time complexity from O($m$) to O($\log r$) for all \op{passes}
operations. Logically the set of $route interval$s should be a sorted \dt{gline}
value but the \secondo{} DBMS restricts us to use a sorted set of $route interval$s
instead.

>>>>

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{2.9.StableNewFlob}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past several data models for the representation of histories of  spatio-temporal data
objects have been developed. Among others we can categorize these data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe the translation
from the \bmodb{} into the network constrained data model and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitudes.
%The very good results of the network data model in this comparison encourages us
%to propose a extension of the \bmodb{} with a additional query set BerlinMOD/Net.
%This new query set covers the specialised challenges for network constrained
%data models and enables us to use the \bmodb{} to compare the capabilities of
%different network data models with respect to this specialised network challenges.
\end{abstract}
\section{Introduction}
In the past several data models for the representation of spatial and
spatio-temporal data objects have been developed. Among others we can categorize
them into data models for objects moving freely in two dimensional space and
data models for network constrained moving objects. For both categories
several different data models have been presented like
\cite{chenzaniolosqlst,335426} for spatio-temporal data objects moving freely in
two dimensional space and \cite{1146465,956692,VazWolfNetMod} for
spatio-temporal data objects that are constrained by given networks,
to name just a few. Objects which are restricted to use existing networks, like
cars are restricted to use road networks, can be represented as moving point object
in both data
models, while objects, which are not restricted by a given network, like people,
can be represented as moving point object only in a data models of free movement
in space.

But if everything can be represented by a data model of free movement
space why do we spend time on network data models? Now, our experiments
presented in this paper show that the network data model outperforms the data model of
free movement in two dimensional space significantly. The network data model uses
less than 60\% of the storage space of the data model of free movement in space, mostly caused
by the fact that the number of units for the moving point objects in the network
data model is less than 50\% of the number of units in the data model of free
movement in two dimensional space. The total run time of the \bmodb{} for the
network data model is less than 50\% of the total run time of the data model of
free movement in two dimensional space. We think that this results show that
it is useful to develop specialised data models for specialised data structures
like network data models for network constrained moving objects to save storage
space and reduce query run times.

For this paper we choosed two representatives one for each data model category.
Both representatives are available in the \secondo{} DBMS and use the same
temporal representation. So we can exclude that different DBMS or temporal
representation issues bias the results of our data model comparison.

The network constrained data model is represented by
the network data model presented in \cite{1146465} and the data model of free
movement in two dimensional space is represented by the data model presented in
\cite{335426}.

For us it seems logical to use the \bmodb{} \cite{BerlinMODVLDB} which is also
available in the \secondo{} DBMS to compare the capabilities of the
both data models. The \bmodb{} is best to our knowledge the first benchmark for
complete spatio-temporal database systems and it is developed in the \secondo{}
DBMS. Furthermore the data generated by the \bmodb{} data generator is restricted
to the streets of the German capital Berlin, such that it can be translated into
a network constrained environment. And not at last the data model used in the
\bmodb{} is the data model of free movement in two dimensional space that we use
for our comparison. So we only have to translate the spatial and spatio-temporal
data types of the \bmodb{} once into our network data model representation.
This simplifies the control of the query results and avoids errors caused by
translation errors.

The translation of the spatial and spatio-temporal data types of the \bmodb{} data
into the network data model representation described in this paper can be seen as
example for the usage of the \bmodb{} with other compatible data representations or
DBMS.

@@

The rest of the paper is organised as follows: In section \ref{sec:relWork} we give a
short reminder of the underlying \secondo{} DBMS (\ref{sec:secondo}), the \bmodb{}
(\ref{sec:bmodb}) and the two data models (\ref{sec:bmodbdatamod}
and \ref{sec:netdatamod}) we choose as representatives for our comparison.
The translation of the \bmodb{} data and query set in the network data model
representation is described in section \ref{sec:Translation}. The resulting
experimental benchmark setup is described in section \ref{sec:scenario} followed
by the results of our experiments in section \ref{sec:results}.
%At least we present our proposed extension of the \bmodb{} for network data models in
%section \ref{sec:newqueries}.
We conclude our work in section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
In the past other data models for free movement in the space
\cite{chenzaniolosqlst,335426} and for network constrained movement have been
presented \cite{1146465,956692,VazWolfNetMod}. As well there are some more
benchmarks \cite{COSTBenchmark, QueriesTheodoridis} and database systems for
spatial and spatio-temporal data types \cite{1054151,HERMES}. But best to our
knowledge they don't actually provide a combination of different implemented
and supported data models together with a existing benchmark feasible for moving
objects in space and network constrained moving objects like the actual
\secondo{} DBMS. So it is self-evident for us to use the \secondo{} DBMS in
combination with the provided data models and the \bmodb{} to compare the
capabilities of the network constrained data model and the data model of free
movement in two dimensional space provided with the \secondo{} DBMS.

In the next subsections we give short reminders of the \secondo{} database system
(\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}), and the both data models
(\ref{sec:bmodbdatamod}, \ref{sec:netdatamod}) we used in our experiments.
\subsection{Secondo}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{686903,1054151} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce  new data types and operations on
this data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the user may publish his extensions
as \secondo{} plugin such that other users can use this plugin to extend their own
\secondo{} system to use the new provided functionalities or repeat the published experiments.
\secondo{} is free available in the web \cite{secondoweb} and comes with a number
of already implemented spatial and spatio-temporal data types
and operations including the spatio-temporal data model of free movement in the
two dimensional space \ref{sec:bmodbdatamod} and the network data model
\ref{sec:netdatamod} used in this paper. Furthermore the \bmodb{} described in
\ref{sec:bmodb} has been developed in the \secondo{} DBMS. For our experiments
we used the \secondo{} version \secver{}.
\subsection{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was presented in \cite{BerlinMODVLDB} \nocite{BerlinMOD} and the
provided scripts for the data generator are implemented as \secondo{} DBMS operations.
The \bmodb{} is available in the web \cite{berlinmodweb} and provides a well defined
data-set and queries for the experimental evaluation of the capabilities of
spatial and spatio-temporal database systems dealing with histories of moving
objects. The \bmodb{} emphasises the development of complete systems
and simplifies experimental repeatability pointing out the weakness and the potency's
of the benchmarked systems.

The data-sets of the \bmodb{} are created using the street map of the German
capital Berlin \cite{bbike} and statistical data about the regions of Berlin
\cite{bevberlin,berlinstadtatlas} as input relations.
The created moving objects represent cars driving in the streets of Berlin,
simulating the behaviour of people living and working in Berlin.
Every moving object has a home node and a work node. Every weekday each car will
do a trip from the home node to work the work node in the morning and vice versa
in the late afternoon. Beside this randomly chosen cars will make additional
trips in the evening and up to six times at the weekend to randomly chosen
targets in Berlin and back home. The \bmodb{} uses the data model of free movement in two
dimensional space described in section \ref{sec:bmodbdatamod}. Because the \bmodb{}
generates all data sets restricted to the street map of Berlin the \bmodb{} can
also be used for network constrained data models, if the spatial and spatio-temporal
data types are translated into a corresponding network data model, like we did
for our experiments.

The number of observed cars and the duration of the observation period can be
influenced by the user setting the $scalefactor$ to different values in the data
generation script of the \bmodb{}. For example at $scalefactor$ 1.0 the data generator
creates 2000 moving point objects observed for 28 days. Each of them sending a
GPS-signal every 2 seconds. This simulated signals are simplified such that time
intervals when a car doesn't move or moves in the same direction at the same
speed are merged into one single time interval. For example: If a car parks in front
of the work node for 8 hours there will be only one entry in the history of the
cars movement with a time interval of 8 hours instead of 14.400 entries
one for each GPS time interval.

@@

In the TBA we have two relations $dataMcar$ and $dataMtrip$. $dataMcar$ contains
the static data for each object like $licence$, $type$, and $model$ together with
an object identifier $moid$. $dataMtrip$ contains for each $moid$ several tuple
each of them containing all units of a single trip of the moving object or a
single unit for a longer stop. For example each time the car drives from home node
to work node is a single trip and each time the car parks in front of the office
is also a single trip.

Beside the moving point objects the \bmodb{} provides several data sets each of them
containing 100 pseudo randomly generated data objects which are used in the
benchmark queries. Table \ref{tab:queryobjects} gives an overview of this query
objects. The \bmodb{} deals also with subsets from this query object sets consisting
of the first or second 10 query objects of a query object set. They are labeled
by the name of the query object set followed by a 1 for the first 10 or a 2 for
the second 10 query objects of the query object set.

\begin{table}[H]
  \begin{tabularx}{1.0\textwidth}{|l|X|}
    \hline
    \textbf{Name of Data Set}&\textbf{Tuple Content}\\
    \hline
    $QueryPoints$&Object identifier and \dt{point} values.\\
    \hline
    $QueryRegions$&Object identifier and \dt{region} value.\\
    \hline
    $QueryInstants$&Object identifier and time stamps.\\
    \hline
    $QueryPeriods$&Object identifier and space of time.\\
    \hline
    $QueryLicences$& Object identifier and a \dt{string} representing a licence value.\\
    \hline
  \end{tabularx}
  \caption{Query Object Relations of \bmodb{}}
  \label{tab:queryobjects}
\end{table}

@@

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item known or unknown object identity,
  \item standard, spatial, temporal, or spatio-temporal dimension,
  \item point, range, or unbounded query interval,
  \item single object or object relations condition type,
  \item with or without aggregation.
\end{itemize}
We will present the 17 queries in more detail in section \ref{sec:queries}
together with our network data model algorithms for this queries.
\subsection{Data Model of Free Movement in two dimensional space}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the same data model of free movement in
two dimensional space presented in \cite{594784,335426,352963}. Spatial
positions are assumed to be located into a two dimensional space. A single spatial
position is represented by the data type \dt{point}. A \dt{point} consists of a pair of
\dt{real} values interpreted as x,y-coordinates in the assumed two dimensional plane.

Streets are represented by \dt{line} values. A \dt{line} value consists of a set
of $half segments$ representing the geometry of the line in the two dimensional space.
Each $half segment$ consists of two \dt{point} values and a Boolean flag telling us
if the left or the right point is the dominating point of the $half segment$.

@@

A car may be represented by a \dt{moving}(\dt{point}) short \dt{mpoint}.
A \dt{mpoint} consist of a set of units called
\dt{unit}(\dt{point}) (short form \dt{upoint}). Each \dt{upoint} consists of a time
interval and two \dt{point} values. The first \dt{point} value represents the
position of the \dt{mpoint} at the start of the time interval and the second
\dt{point} value represents the position of the \dt{mpoint} at the end of the
time interval. It is assumed that the object represented by the \dt{mpoint}
moves on the straight line between this two points with constant speed within the
given time interval. The velocity of the object is given by the ratio from the
distance of the two points and the length of the time interval of the unit.
All units of a \dt{mpoint} must have disjoint time intervals, because a car
cannot be at two different positions at the same time.
The units are sorted by ascending time intervals.

@@

Some other data types of \secondo{} which are used in the \bmodb{} are shown in
table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{1.0\textwidth}{|l|X|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type.\\
\hline
\dt{int} & Usual integer number.\\
\hline
\dt{real} & Usual real number.\\
\hline
\dt{instant} & A point in time.\\
\hline
\dt{periods} & A set of disjoint and not connected spaces of time.\\
\hline
\dt{mbool} & A time dependent boolean value, which will be constant \true{} or \false{}
within each \dt{ubool} \\
\hline
\dt{mreal} & Time dependent real number. Each unit will be defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\subsection{Network Data Model}
\label{sec:netdatamod}
The central idea of the network data model presented in \cite{1146465} is that
every movement is constrained by a given network and every position can be described
relative to this network. The data type \dt{network} is the central
data type in the network data model. All other data types of the network data model
are related to a given \dt{network} object by the unique network identifier that
is part of each \dt{network} object.

The \dt{network} object contains all spatial information of the represented network.
The \dt{network} consists of three main relations ($routes$, $junctions$, and $sections$),
two arrays providing fast access to adjacent network sections,
and some B-Tree and R-Tree indexes to supporting faster access to the main relations.

The relation $routes$ of the \dt{network} contains the attributes of the
streets like $id$, $route curve$, $route length$, and two Boolean flags. The first
flag indicates if the route starts at the lexicographically smaller end point or not.
The second flag indicates if the lanes of the street are separated like on German
Highways or not.

The $junctions$ relation of the \dt{network} contains all attributes of the street
crossings like the two route identifiers of the first and second street meeting at the
crossing, the distance of the crossing from the start of the first respectively
second street, tuple identifiers of the both streets in the routes relation,
tuple identifiers of the sections connected by this junction in the sections relation,
and a connectivity code telling us which lanes of the two streets are connected
by the crossing.

The $sections$ relation of the \dt{network} object contains the
attributes of the street parts between two crossings or a crossing and
the end of the street. This are the route identifier of the street the section
belongs to, the tuple identifier of this street in the routes relation, the start
position and end position of the section on the street, the section curve, and
again two Boolean flags with the same meaning as in the routes relation.

@@

The data type \dt{gpoint} represents single positions in a given network. Besides the
network identifier a \dt{gpoint} consists of a route identifier, a distance from
the start of the route to the position of the \dt{gpoint} and a $side$ value
({$up$, $down$, $none$) telling us if the position is reachable from the $up$
or the $down$ side of the route in case of separated lanes. For simple
streets or positions which are reachable from both sides of the route the side
value is always $none$.

Parts of the network, regardless if they represent paths or regions, are given as
\dt{gline} values. Besides the network identifier a \dt{gline} consists of a set
of $route interval$s, and two Boolean flags. The Boolean flags tell us if the
\dt{gline} is defined and if the set of $route interval$s is sorted.

Each $route interval$ consists of a route identifier identifying the route
the route interval belongs to, and the start and the end position from the
route interval on this route
\footnote{In the original paper the $route interval$ includes a $side$ value
analogous to the \dt{gpoint}. But this parameter is not part of the implementation yet.}.
We call a \label{sec:sortedgline} set of $route interval$s sorted if the following
conditions are fullfilled:
\begin{itemize}
	\item all $route interval$s are disjoint
	\item the $route interval$s are stored in ascending order of their route identifiers
	\item if two disjoint $route interval$s have the same route identifier the
$route interval$ with the smaller start position is stored first
	\item for all $route interval$s the $start position \le end position$
\end{itemize}
Many algorithms take profit from sorted \dt{gline} values. For example: If $n$
is the number of the route intervals in a \dt{gline} the
decision, if a \dt{gpoint} is inside the \dt{gline} needs O($n$) time for unsorted
and O($\log n$) time for sorted \dt{gline} values,

Unfortunately not all \dt{gline} values can be stored sorted. If a \dt{gline}
value represents a path between two \dt{gpoint} in the network, we need the
route intervals exactly in the sequence they are used in the path. This will
nearly never be a sorted set like defined before.  We store \dt{gline}
values sorted whenever this is possible to support faster query execution, and
introduced a Boolean sorted flag. Every algorithm which deals with \dt{gline} values
checks this flag and uses the corresponding code.

@@

In deviation from the original network data model we extended the implementation
of the \dt{mgpoint} with four additional attributes to support faster query execution:
\begin{enumerate}
	\item The total driven distance
	\item A sorted set of $route interval$s representing the positions ever
traversed by the \dt{mgpoint}
	\item A Boolean defined flag for the set of $route intervals$
	\item A spatio-temporal minimum bounding box
\end{enumerate}
The sorted set of $route interval$s was introduced, because analogous to sorted
\dt{gline} values it makes it much faster to decide if a \dt{mgpoint} ever passed
a given network position or not. Instead of a linear check of all $m$ \dt{ugpoint}s
of a \dt{mgpoint} we can perform a binary scan on the much lower number $r$ of the passed $route interval$s.
This reduces the time complexity from O($m$) to O($\log r$) for the \op{passes}
operation. Logically the set of $route interval$s should be a sorted \dt{gline}
value but the \secondo{} DBMS restricts us to use a sorted set of $route interval$s
instead.
